Statistical Inference for inter-arrival times of extreme events in bursty time series

### 2020-07-30 article_springer/article_springer.Rmd
Adjust overview part in the introduction to new strucure

We review the marked renewal process in Section 2, and 
derive a scaling limit theorem for inter-exceedance times in Section 3.
We give a statistical procedure to estimate model parameters via stability plots 
in Section 5, but to set the stage we first discuss inference for the 
Mittag-Leffler distribution in Section 4. A simulation study of the effectiveness of our statistical procedure is given in Section 6. In Section 7 we apply our method to a real data set.  In Section 8, we discuss the memory property of the  Mittag-Leffler distribution, and how it affects the predictive distribution for the time until the next threshold crossing event. Finally we close with a discussion and conclusion in Section 9.
For all statistical computations we have used R [@R]. 
All code and data used for the analysis in this article has been organized into 
an R package `CTRE` (<https://github.com/strakaps/CTRE>).
The source code for the figures generated in this manuscript is 
available online at <https://github.com/strakaps/bursty-POT>.

@@

We assume a dataset of the form $\{(T_i, X_i), i \in I\}$,
where $T_i$ are timestamps and $X_i$ are magnitudes.
In this article, we focus on modelling the distribution of threshold
*exceedance times* rather than the *exceedances* themselves. For the latter, 
see the POT approach [@smith1984threshold;
@leadbetter1991basis; @davison1990models; @beirlantBook Ch.5.3;
@embrechts2013modelling Ch. 6.5.; @ColesBook Ch.4.].
With the previous section setting the stage, we assume that only the large
magnitudes (aka "shocks") follow a MRP.  The smaller magnitudes are possibly
nuisance data that may be irrelevant with respect to the modelling of extremes
and their waiting times. 
Accordingly, we assume a threshold $\ell$, and discard all
data $(T_i, X_i)$ where $X_i \le \ell$, yielding the thinned dataset
$\{(T_i, X_i), i \in I(\ell)\}$, where $I(\ell) \subseteq I$.

@@

For inter-exceedance *times* (IETs), the same idea applies. 
By the Theorem, the inter-exceedance times converge to a Mittag-Leffler
distribution (MLD), and importantly, the tail parameter $\beta$ is 
*independent* of (or *stable* with respect to) the choice of threshold.
For high thresholds, the IET distribution may fit an MLD well, but few
IETs may be left for a fit (high variance). 
For low thresholds, their distribution may be not well represented
by an MLD; or, too many events may be interpreted as 
shocks, resulting in a biased estimate of the parameters of the limiting MLD.
For illustration, consider the case where measurements are taken daily:
Then at the lowest threshold, all IETs equal one day,
leaving no hint at a power-law distribution whatsoever. Raising the threshold
may or may not produce IETs that start exhibiting a power-law; and both cases
are represented by the MLD, with $\beta < 1$ or $\beta = 1$.

Suppose the "correct" threshold $\ell_0$ was known, in the sense that 
thresholding at $\ell_0$ would result precisely in an underlying MRP 
$\{(W_i, J_i): i \in I\}$ as defined in Section \ref{sec:CTRE}. 
In this case, we could apply any tail (and scale) estimator such as e.g.
Hill's estimator [@hill1975simple]
to the $W_i$ and thus infer the distribution of 
inter-exceedance times.

But guessing the correct threshold $\ell_0$ and proceeding with this
type of inference is less favourable to modelling IETs with the MLD, 
as we now explain.
In real-world datasets, a "correct" threshold may not exist: the data 
$\{(T_i, X_i): i \in I\}$ may be a mixture of an MRP with some other
time series $\{(T'_i, X'_i): i \in I'\}$, and the supports $\mathcal S'$
and $\mathcal S$ of the distributions
of $X'_i$ and the magnitudes $J_i$ (aka the MRP marks) may overlap.
Then thresholding at e.g. $\ell_0 = \sup \mathcal S'$ 
no longer results in a "clean" MRP: some $J_i$'s will be removed, meaning
that some $W_i$'s will be merged into a longer IET.
The more the threshold $\ell_0$ cuts into $\mathcal S$ from below, the more
the distributions of IETs will resemble a MLD, according to the Theorem.
As we do not know to what extent this effect occurs, 
and since the limit is an MLD, it is practicable to fit an MLD to the IETs to 
begin with.

@@

## Weighing the evidence for heavy tails

>>>>

We review the marked renewal process in Section 2.1, and 
derive a scaling limit theorem for inter-exceedance times in Section 2.2.
We give a statistical procedure to estimate model parameters via stability plots in Section 3.1 and 3.4, but to set the stage we first discuss inference for the Mittag-Leffler distribution in Section 3.2 as well as a Likelihood-ratio test to assess the goodness of fit. A simulation study of the effectiveness of our statistical procedure is given in Section 4. In Section 5 we apply our method to a real data set.  In Section 6, we discuss the memory property of the  Mittag-Leffler distribution, and how it affects the predictive distribution for the time until the next threshold crossing event. Finally we close with a discussion and conclusion in Section 7.
For all statistical computations we have used R [@R]. 
All code and data used for the analysis in this article has been organized into an R package `CTRE` (<https://github.com/strakaps/CTRE>).
The source code for the figures generated in this manuscript is 
available online at <https://github.com/strakaps/bursty-POT>.

@@

We assume a dataset of the form $\{(t_i, x_i), i \in I\}$,
where $t_i$ are timestamps and $x_i$ are magnitudes.
In this article, we focus on modelling the distribution of threshold
*exceedance times* rather than the *exceedances* themselves. For the latter, 
see the POT approach [@smith1984threshold;
@leadbetter1991basis; @davison1990models; @beirlantBook Ch.5.3;
@embrechts2013modelling Ch. 6.5.; @ColesBook Ch.4.].
With the previous section setting the stage, we assume that only the large
magnitudes (aka "shocks") follow a MRP.  The smaller magnitudes are possibly
nuisance data that may be irrelevant with respect to the modelling of extremes
and their waiting times. 
Accordingly, we assume a threshold $\ell$, and discard all
data where $X_i \le \ell$, yielding the thinned dataset
$\{(t^{(l)}_i, x^{(l)}_i), i \in I(\ell)\}$, where $I(\ell) \subseteq I$.

@@

For inter-exceedance *times* (IETs), the same idea applies. 
By the Theorem, the inter-exceedance times converge to a Mittag-Leffler
distribution (MLD), and importantly, the tail parameter $\beta$ is 
*independent* of (or *stable* with respect to) the choice of threshold.
For high thresholds, the IET distribution may fit an MLD well, but few
IETs may be left for a fit (high variance). 
For low thresholds, their distribution may be not well represented
by an MLD; or, too many events may be interpreted as 
shocks, resulting in a biased estimate of the parameters of the limiting MLD.
<!-- For illustration, consider the case where measurements are taken daily: -->
<!-- Then at the lowest threshold, all IETs equal one day, -->
<!-- leaving no hint at a power-law distribution whatsoever. Raising the threshold -->
<!-- may or may not produce IETs that start exhibiting a power-law; and both cases -->
<!-- are represented by the MLD, with $\beta < 1$ or $\beta = 1$. -->

<!-- Suppose the "correct" threshold $\ell_0$ was known, in the sense that  -->
<!-- thresholding at $\ell_0$ would result precisely in an underlying MRP  -->
<!-- $\{(W_i, J_i): i \in I\}$ as defined in Section \ref{sec:CTRE}.  -->
In case we would have data of a MRP $\{(W_i, J_i): i \in I\}$ as defined in Section \ref{sec:CTRE}, we could apply any tail (and scale) estimator such as e.g. Hill's estimator [@hill1975simple] to the $W_i$ and thus infer the distribution of inter-exceedance times. However, in real-world datasets, the MRP assumption is unrealistic for low thresholds. The unknown data generating process is likely to be more complex, with dependencies or mixtures of time series at low thresholds, structures that possibly vanish for higher thresholds. Fitting just the IETs means that only the exceedances as well as the times inbetween have to satify the iid assumption.

<!-- But guessing the correct threshold $\ell_0$ and proceeding with this -->
<!-- type of inference is less favourable to modelling IETs with the MLD,  -->
<!-- as we now explain. -->
<!-- In real-world datasets, a "correct" threshold may not exist: the data  -->
<!-- $\{(T_i, X_i): i \in I\}$ may be a mixture of an MRP with some other -->
<!-- time series $\{(T'_i, X'_i): i \in I'\}$, and the supports $\mathcal S'$ -->
<!-- and $\mathcal S$ of the distributions -->
<!-- of $X'_i$ and the magnitudes $J_i$ (aka the MRP marks) may overlap. -->
<!-- Then thresholding at e.g. $\ell_0 = \sup \mathcal S'$  -->
<!-- no longer results in a "clean" MRP: some $J_i$'s will be removed, meaning -->
<!-- that some $W_i$'s will be merged into a longer IET. -->
<!-- The more the threshold $\ell_0$ cuts into $\mathcal S$ from below, the more -->
<!-- the distributions of IETs will resemble a MLD, according to the Theorem. -->
<!-- As we do not know to what extent this effect occurs,  -->
<!-- and since the limit is an MLD, it is practicable to fit an MLD to the IETs to begin with. -->

@@

## Goodness-of-Fit Test
